A Bayesian model of latitudinal temperature gradients applied to the Early Eocene

### 2023-02-09 manuscript/discards.rmd
discussion eco prox

>>>>

## Discussion discards

or even adjacent time intervals 
the fossil record indicates that many taxa

 This approach is showcased by nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions, which work on the premise that the climatic requirements of ancient plant taxa were similar to those of there nearest modern relatives [@Kvacek2007]. A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...]. Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantitative modelling framework. Here, we employ a similar approach, constructing probability distributions broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.

Quantifying climatic constraints provided by fossil organisms relies on the assumption that the climatic conditions under which they lived are similar to those occupied by their modern relatives.

 Similarly

Our approach for quantifying climatic constraints as normal distributions broadely based on their modern, climatic equivalents, deviated from recent

tQuantifying climatic constraints provided by fossil organisms

relies on the assumption that the climatic conditions under which they lived are similar to those occupied by their modern relatives. Similarly

Whilst our approach provides normal probability distributions for deriving temperature estimates of corals and mangrove communities,

As with other nearest-living relative methods, our ecological temperature estimates come with limitations: 1)

Although the quantification of each climate constraint needs to account for specific information, o

However, our simple approach to constructing probability distributions for the thermal range of coral reefs has limitations. The use of only modern coral reefs and mangrove taxa as analogues for their fossil counterparts assumes that the thermal preferences of these taxa have remained unchanged over time, which may not always be the case. In addition, the use of only nearest living relatives as proxies assumes that past reefs may have thrived under the same temperatures as those observed today, which may not always be accurate. Climate change and shifting biogeographical patterns may have resulted in different thermal conditions for ancient reefs. Therefore, our approach should be supplemented with other methods that take into account changes in the thermal tolerance of reef taxa over time.

Ecological data and other semi-quantiative or qualitative climate constraints are readily incorporated in this modelling framework: The hierarchical structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. Existing palaeoclimate studies have mostly

### 2023-02-09 manuscript/manuscript.Rmd
discussion eco prox

Our results show that incorporating quantified ecological temperature constraints can improve temperature reconstructions base on geochemical proxies alone. Given the significant body of research on the climatic significance of e.g. plants [@ @ @], reptiles [@ ] and lithology types [@ ], there is great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Although the quantification of each climate constraint needs to account for specific information, our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.

As with other nearest-living relative methods, our ecological temperature estimates come with limitations: 1)

However, our simple approach to constructing probability distributions for the thermal range of coral reefs has limitations. The use of only modern coral reefs and mangrove taxa as analogues for their fossil counterparts assumes that the thermal preferences of these taxa have remained unchanged over time, which may not always be the case. In addition, the use of only nearest living relatives as proxies assumes that past reefs may have thrived under the same temperatures as those observed today, which may not always be accurate. Climate change and shifting biogeographical patterns may have resulted in different thermal conditions for ancient reefs. Therefore, our approach should be supplemented with other methods that take into account changes in the thermal tolerance of reef taxa over time.

Ecological data and other semi-quantiative or qualitative climate constraints are readily incorporated in this modelling framework: The hierarchical structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. Existing palaeoclimate studies have mostly

>>>>

Our results show that incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone. Climate-sensitive plant communities [@ @ @], reptiles [@ ], leaf shapes [@], and lithologies [@ ] offer great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic. Our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Faucette2007; @Kvacek2007]. These methods have the problem that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@]. Although coral reefs are threatened by warming sea surface temperatures today [@], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012]. Tropical temperatures predicted by the geochemical record indicate hotter-than-modern tropical temperatures (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2). Future studies may try to resolve this dilemma by applying machine learning approaches to inferring quanitative temperatures from ecological and lithological proxies (e.g. @Salonen2019; @Wei2020; @Chandra2021) in novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures. This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking a good coverage of geochemical proxies via their fossil or sedimentary record.
